Edge sign prediction in general graphs and Correlation Clustering

### 2017-10-27 edge/signed.tex
2-CC version of the connection between edge and signed
The previous approaches was indeed not correct, for reasons explained
in the aside paragraph :(

Here we show why \autoref{p:edge_full} can model signed graphs as a special case, namely with $k=2$
and specifically crafted profiles.
Remember the learning
bias we introduced in the previous chapter \vpageref{text:cc_new_bias}: every node belongs to
exactly one of the $K$ clusters available, is connected positively with the other nodes belonging to
the same cluster, and is connected negatively with the nodes belonging to every other clusters. Now
let the dimension of profiles be $d=K$ and the number of directions $k=2$. Moreover, we define the
two direction vectors as $w_1 = \frac{\onev}{\sqrt{d}} = w_+$ and $w_2 = -w_+ = w_-$, where
$\onev$ is the vector of dimension $d$ with only $1$ coordinates. Finally, for a node $u$ in
cluster $i$, we assume its profile $x_u$ is set as follow: all component are equal to $-b$ except
for the \ith{} one, which is equal to $a$, where $a$ and $b$ are real constants we define now.

First, because $x_u$ is a unit vector, we have that $a^2 + (K-1)b^2=1$, from which $a =
\sqrt{1-(K-1)b^2}$ whenever $1-(K-1)b^2 \geq 0$, or equivalently $b \leq \frac{1}{\sqrt{K-1}}$.
Next, we look at $\left(x_u \circ x_v\right)^T \onev = {s_{uv}}^T \onev = \frac{1}{\sqrt{d}}
\sum_{i=1}^d  {s_{uv}}_{;i}$ for any edge $(u,v) \in E$. If $u$ and $v$ are in the same cluster, then
$\sum_{i=1}^d  {s_{uv}}_{;i} = a^2 + (K-1)b^2 = 1$, implying that ${s_{uv}}^T w_+ = \frac{1}{d} >
{s_{uv}}^T w_- = -\frac{1}{d}$. Otherwise, $\sum_{i=1}^d  {s_{uv}}_{;i} = (K-2)b^2 - 2ab = (K-2)b^2 -
2b\sqrt{1-(K-1)b^2} = f_K(b)$. By computing the derivative of $f_K$ with respect to $b$, we find
that it reaches a minimum of $-\frac{1}{K-1}$ at $b = \frac{1}{\sqrt{K(K-1)}} \leq
\frac{1}{\sqrt{K-1}}$. In this case, ${s_{uv}}^T w_+ = -\frac{1}{(K-1) \sqrt{d}} < {s_{uv}}^T w_- =
\frac{1}{(K-1) \sqrt{d}}$. To maximize $\sum_{(u,v)\in E} \max_{\ell \in \{1, 2\}} {s_{uv}}^T
w_\ell$, we therefore assign $w_+$ to edges whose both endpoints belong to the same cluster, and
$w_-$ to edges joining two different clusters.

Moreover, given this assignment, we can show that $w_+$ and $w_-$ are the optimal directions, under
the additional uniformity assumption that there are exactly $p$ edges within every cluster, and $q$
edges between any two clusters. First, we define
\begin{equation*}
  s_{\mathrm{inner}} = \sum_{(u,v) \in E;\, \cluster(u)=\cluster(v)} s_{uv} \qquad \text{and} \qquad
  s_{\mathrm{outer}} = \sum_{(u,v) \in E;\, \cluster(u)\neq \cluster(v)} s_{uv}
\end{equation*}
to be the sum of all edge vectors within and between clusters respectively. By our choice of $a$ and
$b$, we have that $s_{\mathrm{inner}} = p\onev$ and $s_{\mathrm{outer}} = -\frac{q}{2}\onev$. Now
recall that the solution of $\argmax_{w\in \dsphere} w^T c$ is $\frac{c}{||c||}$. Therefore, the
best direction to assign to inner edges is indeed $w_+$, and so is $w_-$ to outer edges.

\begin{aside}
For inner edge vectors, $s_{uv}$ in cluster $i$ has $a^2$ as its \ith{} component, and $b^2$
elsewhere. There are $p$ of them in cluster $i$, so summing over all clusters we get
$s_{\mathrm{inner}} = p\left(a^2 + (K-1)b^2\right)\onev = p\onev$.
For outer edge vectors, we first consider the $q$ edges between $C_1$ and $C_2$,
whose $s_{uv}$ is
$\begin{pmatrix}
  -ab \\
  -ab \\
  b^2 \\
  b^2 \\
  \vdots \\  
\end{pmatrix}$. Likewise, between $C_1$ and $C_3$ we have
$s_{uv}=\begin{pmatrix}
  -ab \\
  b^2 \\
  -ab \\
  b^2 \\
  \vdots \\  
\end{pmatrix}$. We thus we see that the first component is always $-ab$, that there is another
such term at the other cluster index, while the rest is $b^2$. Summing all the vectors incident to
$C_1$, we get
$q\begin{pmatrix}
  -ab(K-1) \\
  b^2(K-2) -ab \\
  b^2(K-2) -ab \\
  \vdots \\  
\end{pmatrix}$. Similarly, the sum of all vectors incident to $C_2$ is
$q\begin{pmatrix}
  b^2(K-2) -ab \\
  -ab(K-1) \\
  b^2(K-2) -ab \\
  \vdots \\  
\end{pmatrix}$, and so on for every cluster $i$. Summing over all clusters and dividing by $2$ to
avoid double counting, we get $s_{\mathrm{outer}} = \frac{q}{2}\left( (K-1)\left(b^2(K-2) -ab\right)
-ab(K-1) \right)\onev = \frac{q(K-1)}{2}\left(b^2(K-2) -2ab\right)\onev =
\frac{q(K-1)}{2}f_K(b)\onev = -\frac{q(K-1)}{2} \frac{1}{K-1} \onev = -\frac{q}{2}\onev$.
\end{aside}

Granted, the situation where all the nodes of a cluster have exactly the same profile and where the
connections are so regular is highly idealized. Moreover, we did not formally prove that there is
not another assignment (with different direction vectors) that would give a better objective
value\footnote{Although given the symmetry of the problem, that would be surprising.}. Our point is
rather that profiles and directions can be seen as a \enquote{latent} explanation for the signs in
a balanced signed graphs, like our generative model of the first chapter.

>>>>

The learning bias of \autoref{chap:cc} implicitly defined an idealized \esp{} problem as:
\begin{problem}[idealized \esp{}]
  \label{p:edge_esp}
  Given $n$ nodes belonging latently to $K$ groups, predict whether two nodes $u$ and $v$ belong to
  the same group (that is $(u,v)$ is positive) or not (that is $(u,v)$ is negative).
\end{problem}
In \autoref{sec:gtx} we study in particular the important special case where $K=2$. Now consider an
instance of \autoref{p:edge_full} where $k=2$, and where the profiles and edges are based on group
membership. We show that the optimal solution of such an instance is exactly the solution of this
idealized \esp{} problem on the same graph.

Let us first define this instance formally. Given a graph $G=(V,E)$, we partition the nodes in two
groups $L$ and $R$. We then define the nodes profiles $X$ as follow. We first draw \uar{} a vector $p$
from $\{-1, 1\}^d$, then let the profiles in $L$ be $\frac{p}{\sqrt{d}}$ and the
profiles in $R$ be $-\frac{p}{\sqrt{d}}$. We call this instance $\mathcal{I} = (G, X, k=2)$. With
this choice, we have
\begin{equation*}
  s_{uv} =
  \begin{cases}
    \frac{\onev}{d} = s_{\mathrm{inner}} & \quad \text{if $u$ and $v$ are in the same group,} \\
    -\frac{\onev}{d} = s_{\mathrm{outer}} & \quad \text{if $u$ and $v$ are in different groups.}
  \end{cases}
\end{equation*}
Let us denote by $m_i$ the number of edges that are internal to $L$ and $R$, and by $m_o = |E| -
m_i$ the number of edges between $L$ and $R$.

\begin{prop}
  Let $\mathcal{D}_2 = \{w_1^\star, w_2^\star\}$ be an optimal solution of \autoref{p:edge_full} on the
  instance $\mathcal{I}$. Then $\mathcal{D}_2$ provides the following optimal solution to
  \autoref{p:edge_esp}: each every edge explained by $w_1^\star$ is predicted positive, and every
  edge explained by $w_2^\star$ is predicted negative.
\end{prop}
\begin{proof}
  Take any $w_1, w_2 \in \dsphere \times \dsphere$ such that $w_1 \neq w_2$. Assume without loss of
  generality that $w_1 = \argmax_{w\in \{w_1, w_2\}} {s_{\mathrm{inner}}}^T w$. This implies that:
  \begin{equation*}
    {s_{\mathrm{inner}}}^T w_1 = \frac{1}{d}\sum_{i=1}^d {w_1}_{;i} >
    \frac{1}{d}\sum_{i=1}^d {w_2}_{;i} = {s_{\mathrm{inner}}}^T w_2 \,.
  \end{equation*}
  In turn, we have that:
  \begin{equation*}
    {s_{\mathrm{outer}}}^T w_2 = -\frac{1}{d}\sum_{i=1}^d {w_2}_{;i} >
    -\frac{1}{d}\sum_{i=1}^d {w_1}_{;i} = {s_{\mathrm{outer}}}^T w_1 \,.
  \end{equation*}
  We can thus rewrite the maximization objective \eqref{eq:edge_full} of \autoref{p:edge_full} as:
  \begin{align}
    \label{eq:edge_special}
    \argmax_{w_1,w_2 \in \dsphere} & \quad \sum_{u,v \in E} \max_{w \in \{w_1, w_2\}} {s_{uv}}^T w = \\
    \argmax_{w_1,w_2 \in \dsphere} & \quad
    \sum_{\substack{u,v \in E\\ s_{uv} = s_{\mathrm{inner}}}} {s_{\mathrm{inner}}}^T w_1 +
    \sum_{\substack{u,v \in E\\ s_{uv} = s_{\mathrm{outer}}}} {s_{\mathrm{outer}}}^T w_2 = \notag \\
    \argmax_{w_1,w_2 \in \dsphere} & \quad m_i \frac{1}{d} \onev + m_o \frac{-1}{d} \onev \notag
  \end{align}
  One can check using Lagrange multipliers that for any $c \in \Rbb^n$, $\max_{w \in \dsphere} c^T w
  = \frac{c}{\|c\|}$. The solution of \eqref{eq:edge_special} is therefore:
  \begin{equation*}
    w_1^\star = \frac{\onev}{\sqrt{d}} \qquad \text{and} \qquad w_2^\star = \frac{-\onev}{\sqrt{d}}
  \end{equation*}
  As claimed, the edges associated with $w_1^\star$ (respectively $w_2^\star$) are the edges that
  are to be predicted positive (respectively negative) in \autoref{p:edge_esp}.
\end{proof}

\begin{aside}
  It is natural to ask how this result can be extended to more than $K=2$ groups. The most natural
  idea is to use one-hot encoding of the group membership as profiles, with additional negative
  values as in the example of \autoref{fig:edge_exe}. Namely, set the profiles of the \ith{} group
  as follow: all component are equal to $-b$ except for the \ith{} one, which is equal to $a$, where
  $a$ and $b$ are real constants such that $a^2 + (K-1)b^2=1$. However, in that case, we check
  numerically that $w_+ = \frac{\onev}{\sqrt{d}}$ and $w_- = \frac{-\onev}{\sqrt{d}}$ is the not the
  optimal solution, and that a solution achieving a larger objective function value does not cluster
  the edges between inner and outer.

  The next thing I tried was to use the approach described later in \autoref{sub:edge_generation} in
  order to optimize the profiles of a complete graphs with $K$ groups, given $w_+$, $w_-$ and the
  inner/outer edges assignment. As one could expect from the symmetry of the problem, all the nodes
  in the same group have the same profile. However, there are no clear pattern in those profile, and
  once again there exists better solutions than $\{w_+, w_-\}$, which partition the edges in two
  roughly balanced sets.
\end{aside}
